### 2010-08-11 Thesis/Chapter4.tex
- Update 11 August -
Added free-form deformation, shape matching and mass-spring systems + big update for bib



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@149 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Conversely, when medical simulation is applied to training, such a precision in terms of distance is not required. The physician would not even visually notice reasonable errors as soon as it looks realistic enough. However, in some medical procedures, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation. For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent the loop formation. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures.

@@

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness.

@@

	\subsection{Mass-spring}

@@

>>>>

Conversely, when medical simulation is applied to training, such a precision in terms of distance is not required. The physician would not even visually notice reasonable errors as soon as it looks realistic enough. However, in some medical procedures, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation. For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent loop formation. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures.

@@

	\subsection{Free-form deformation}
The first needs to deforming solid objects came from for the field of solid modelling. The goal was to allow the designer to shape an object into the form he wishes in a similar way that clay is manipulated by a sculptor's hands. \cite{Sederberg86} introduced a technique called \emph{free-form deformation} based on the paper of \cite{Barr84}. It involves a mapping from $ \field{R}^3 $ to $ \field{R}^3 $ through a tensor product trivariate Bernstein polynomial. The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surfaces patches (see Fig.~{chap4:fig-cubeFFD}). 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}
\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}
\end{center}
\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped. Control points are indicated by small white diamonds. The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points (right).}
\label{chap4:fig-cubeFFD}
\end{figure}
The capability of locally applying deformations makes the technique strongly analogous to sculpting with clay, which was the sought target. Although the authors do not give details on computational efficiency, their work constitutes the beginnings of deformable solid objects.

Coquillard later developed the idea of Extented free-form deformation \citeyearpar{Coquillart90} and Animated free-form technique \citeyearpar{Coquillart91}. The idea of free-form deformation with lattices of arbitrary topology was introduced by \cite{MacCracken96}. Finally, \cite{Schein04} presented a technique to facilitate the incorporation of discontinuities in a model while deforming it.

	\subsection{Shape matching}
One field particulary interested in modelling deformable objects in a very efficient way is the game industry. Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws. Most of the time, we can tolerate a degradation of realism as long as the result looks realistic. \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}. We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules. No connectivity information is required. The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity. The positions in the deformed configuration are noted $ \mathbf{x}_i $. The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $. The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
\begin{equation}
\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
\end{equation}
is minimal. The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively. The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $. The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition. Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
\begin{equation}
\label{chap4:gi}
\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
\end{equation}
Once the goal positions are known, they are used to integrate the positions of the particules:
\begin{equation}  
	\begin{cases} 
		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
	\end{cases}
\end{equation}
where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness. Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter. Linear transformation can represent shear and stretch while quadratic transformation add twist and bending. Of course, this technique is not physically realistic. But this is beyond the point. This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see Fig.~\ref{chap4:fig-shoes}). Another illustration of this stability is shown Fig.~\ref{chap4:fig-duck}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
\end{center}
\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time.}
\label{chap4:fig-shoes}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
\vspace{0.1cm}
\includegraphics[width=11cm]{chapter4/duck2.pdf}
\end{center}
\caption[Squeezing of a duck using Shape Matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique. Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
\label{chap4:fig-duck}
\end{figure}

Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry. From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh. The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions. A particule is placed at each vertex of the lattice and associated with a shape matching region. Their technique of \emph{Fast Lattice Shape Matching} allows them to compute per-region transformations in an efficient manner (see Fig.~\ref{chap4:fig-penguins}).	
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/penguins.pdf}
\end{center}
\caption[Penguins animation with Fast Lattice Shape Matching]{(Left) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching. (Right) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles). The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
\label{chap4:fig-penguins}
\end{figure}

One limitation of fast lattice shape matching is that small features of a geometry yield an explosion of the runtime cost. Indeed, a small surface feature may require fine sampling in order to be deformed independently from non-adjacent material, but this fine sampling must be applied to the whole object. Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling. This adaptative sampling is illustrate with Fig.~\ref{chap4:fig-flowers}. Their technique also handles efficient topological changes.
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
\end{center}
\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers. A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene. However, the models are dynamically refined when touched by the user, as shown in the left image.}
\label{chap4:fig-flowers}
\end{figure}

@@

	\subsection{Mass-spring model}
A classic technique for modelling a deformable object is the mass-spring system. In this method, the geometry is described by a network of masses connected together by springs. The force $ F $ exerted onto each mass by the spring may be computed with:
\begin{equation}
\label{chap4:massSpring1}
F_{\text{spring}} = - k x,
\end{equation}
where $ k $ is the spring stiffness constant, $ x $ represents the stretch of the spring from its rest position. From our experience, we know that a mass attached to a single spring has a tendency to oscillate. If no appropriate measure is taken, a network of springs will a forciori oscillate as well. To prevent the system from oscillating, a damper is also added between each couple of masses. The force created by the damper acts to reduce the velocity and may be expressed as:
\begin{equation}
\label{chap4:massSpring2}
F_{\text{damper}} = - d v
\end{equation}
where $ d $ is the damping constant and $ v $ the derivative of $ x $ with respect to time. The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations. Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{a} $ yields the standard differential equation for a mass-spring system:
\begin{equation}
\label{chap4:massSpring3}
m \ddot x = - d \dot x - k x.
\end{equation}
For complex systems with more than one spring attached to each mass, the right-hand side of the equation above becomes more complex as it needs to be repeated and adjusted for each spring. The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}). Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}). This equation must be numerically solved for each mass, either by explicit or implicit methods. A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.

Mass-spring systems have been extensively used and are still very popular. And this is not surprising because they are easy to implement and computationally efficient. Of course, they have drawbacks. But as we will see shortly, various recent works attempt to circumvent them. One limitation often mentioned is the difficulty to derive spring stiffnesses from elastic properties (Young's modulus and Poisson's ratio). To overcome this deficiency, \cite{Lloyd07} mention two ways of obtaining the parameters of a mass-spring model. The first one consists in varying the parameters until the behaviour of the system is similar to the one obtained through the experiments or the one obtained with the Finite Element Method \citep{Bianchi04}. The second way consists in establishing an analytical reasoning for calculating the constants of the mass-spring model. For instance, starting from the definition of Young's modulus, Poisson's ration, shear and bulk modulus, \cite{Baudet07} apply simple tests to their mass-spring model to find the most appropriate stiffnesses. \cite{Lloyd07} proposes a linearisation of the mass-spring equations with the aim of equating the linearised stiffness matrix to the stiffness matrix of a linear finite element method. By following the same method, \cite{SanVicente09} derived a mass-spring model equivalent to a linear finite element model for maxillofacial surgery simulation.

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in Fig.~\ref{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[h]
\begin{center}
\includegraphics[width=8cm]{chapter4/undesiredAnisotropy.pdf}
\end{center}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (left) shows signs of anisotropy, the hexahedral mesh (right) with springs aligned along the gravity does not.} 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

A mass-spring model may also be enhanced to feature viscoelasticity. Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles. Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a spring-mass system. They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics. \cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery. They tune their parameters by a simple optimisation procedure to fit the mechanical response obtained on a set of experimental data.

	\subsection{Modal analysis}

@@

	\subsection{Point-based}
